<!-- README.md is generated from README.Rmd. Please edit that file -->


```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%",
  dpi=300,fig.width=7,
  fig.keep="all",
  eval=FALSE
)
```

# bigPLSR <img src="man/figures/logo_bigPLSR.svg" align="right" width="200"/>

# bigPLSR, PLS regression models in R
## Frédéric Bertrand and Myriam Maumy

<!-- badges: start -->
[![R-CMD-check](https://github.com/fbertran/bigPLSR/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/fbertran/bigPLSR/actions/workflows/R-CMD-check.yaml)
[![R-hub](https://github.com/fbertran/bigPLSR/actions/workflows/rhub.yaml/badge.svg)](https://github.com/fbertran/bigPLSR/actions/workflows/rhub.yaml)
<!-- badges: end -->


The goal of bigPLSR is provide Partial Least Squares Regression models in a high dimensional setting in R.



Support for parallel computation and GPU is being developed.


This website and these examples were created by F. Bertrand and M. Maumy.

## Installation

You can install the released version of bigPLSR from [CRAN](https://CRAN.R-project.org) with:

```{r, eval = FALSE}
install.packages("bigPLSR")
```

You can install the development version of bigPLSR from [github](https://github.com) with:

```{r, eval = FALSE}
devtools::install_github("fbertran/bigPLSR")
```

# Example


